---
title: "Brazil Analyses-Clean"
output: html_notebook
---


```{r}
# set timezone
Sys.setenv(TZ = "GMT")

library(car) # calls the correct libraries
library(doBy) # a package that will allow you to quickly visualize the data within categories. It allows a bit more control than describeBy within psych 
library(sciplot) # a new library that allows for quick interaction plots
library(ggplot2)
library(DescTools)
library(psych)
library(plyr)
library(lattice)
library(dplyr)
library(emmeans)
library(HH)
library(gvlma)
library(productplots)
library(lubridate)
library(dplyr)
library(MASS)
library(candisc)
library(ICSNP)
library(MANOVA.RM)
library(betareg)
library(vegan)
library(outliers)
library(lme4)
library(reshape2)
library(PMCMRplus)
library(PMCMR)


birbs<-read.csv("Brazil Motus 2019-2020.csv")
#motus csv template 2019-2020 combined

```

Prepping Motus Data:
```{r}
#converting existing columns to time in excel
birbs$Deploy.ts<-as.POSIXct(birbs$Deploy.ts,format='%m/%d/%Y %H:%M')
birbs$Departure.ts<-as.POSIXct(birbs$Departure.ts, format='%m/%d/%Y %H:%M')
birbs$final.ts<-as.POSIXct(birbs$final.ts, format='%m/%d/%Y %H:%M')

```

```{r}
birbs$Size.Adjusted.Mass<-(birbs$weightGrams/(birbs$wingMillimeters^3))*10000

birbs$Duration.Stay<-difftime(birbs$Departure.ts,birbs$Deploy.ts, units='days')

birbs$tot.det<-difftime(birbs$final.ts,birbs$Deploy.ts, units='days')
birbs$tot.det<-as.numeric(birbs$tot.det)
birbs$log.det<-log10(birbs$tot.det)

birbs$north.dur<-difftime(birbs$final.ts,birbs$Departure.ts, units='days')
```


```{r}
hist(as.numeric(birbs$Duration.Stay), breaks=(50), main="Histogram of zduration of Stay")
```


```{r}
#remove NAs and "toss"points (detected <4 days)
birbs.reduced<-dplyr::filter(birbs, Duration.Stay > 4)
birbs.reduced<-dplyr::filter(birbs.reduced, Strategy!="few")
birbs.reduced<-dplyr::filter(birbs.reduced, Strategy!="NA")
birb.test<-dplyr::filter(birbs, Duration.Stay < 4)
#histogram of duration
hist(as.numeric(birbs.reduced$Duration.Stay), breaks=30, main="Histogram of Duration of Stay")

birbs.reduced$utcYearStart<-as.factor(birbs.reduced$utcYearStart)

birbs.reduced<- birbs.reduced %>% dplyr::select(-mfgID,-dateBin,-tagProjectID,-tagDeployID,-period,-utcMonthStart,-utcDayStart,-utcHourStart,-utcMinuteStart,-deferSec,-bandNumber,-markerNumber,-markerType,-billMillimeters,-headMillimeters,-fatScore,-comments,-elevationMeters)
```


count strategies:
```{r}
stratcount<-plyr::count(birbs.strat, c("Strategy","utcYearStart"))
print(stratcount)

remcount<-plyr::count(birb.test, c("utcYearStart"))
print(remcount)
```

means: 
```{r}
mean_and_var <- summaryBy(Duration.Stay ~ utcYearStart + Strategy, data=birbs.strat, FUN=c(mean,var,sd))
mean_and_var 

```


Average time on tidal flat for birds on days they're using the tidal flats
```{r}
avg.tfl<-read.csv("daily.use.csv")
```


```{r}
tfl.summary <-avg.tfl %>%
  group_by(motusTagID,Strategy) %>%
  dplyr::summarise(mean=mean(time.TF.hours), var=var(time.TF.hours))
  
```


```{r}
mean_and_varTFstrat <- summaryBy(mean ~ Strategy, data=tfl.summary, FUN=c(mean,var,sd))
mean_and_varTFstrat 

```

Bill model:
```{r}
bill.fit1<- lm(billMillimeters~Strategy, data=birbs.reduced)

Anova(bill.fit1, type="II") # generates a typical summary of the output.


resid.bill1=resid(bill.fit1) # generating residuals

qqnorm(resid.bill1) # plotting residuals for visual test of normality
qqline(resid.bill1)
shapiro.test(resid.bill1)
hist(resid.bill1)
boxplot(resid.bill1~birbs.reduced$Strategy)

cochran.test(bill.fit1, data = birbs.reduced)

emmeans(bill.fit1, pairwise~Strategy)
```

Bill plot:
```{r}
billplot<-ggplot(aes(x=Strategy, y=billMillimeters), data = birbs.reduced)+
  geom_boxplot() +
   geom_jitter(shape=16, position=position_jitter(0.2)) +
  scale_x_discrete((name = "Movement strategy"),labels=c("Tidal strategy","Sunrise/Sunset strategy","Mixed strategy"))+
  labs(y="Bill length (mm)") +
  theme_classic(base_size = 15) 
  #geom_hline(yintercept=19, color="chartreuse4")
billplot
```



Metabolite Analyses:
```{r}
all.birbs<-read.csv("2019-2020 Brazil Banding.csv")
```

Plasma T analysis:
```{r}

Plasma.T.birbs<-birbs.reduced %>%
  filter(Plasma.Value > 0)

Plasma.T.birbs$log.val<-log10(Plasma.T.birbs$Plasma.Value)

all.birbs$Year<-as.factor(all.birbs$Year)


all.birbs.PT<-all.birbs %>%
  filter(Plasma.T.Val >0)

all.birbs.PT$log.val<-log10(all.birbs.PT$Plasma.T.Val)
```

Tagged bird TRIG model:
```{r}
birbs1PT.fit <- lm(log.val~ Strategy*utcYearStart, data=Plasma.T.birbs) # model for 2-way anova

Anova(birbs1PT.fit, type="II") # generates a typical summary 
emmeans(birbs1PT.fit, pairwise ~ Strategy)

cochran.test(birbs1PT.fit, data = Plasma.T.birbs)

resid.anova.PT=resid(birbs1PT.fit) # generating residuals

qqnorm(resid.anova.PT) # plotting residuals for visual test of normality
qqline(resid.anova.PT)
shapiro.test(resid.anova.PT)
hist(resid.anova.PT)
boxplot(resid.anova.PT~Plasma.T.birbs$Strategy)
```

All bird TRIG model:
```{r}
plasmaTSAM.fit <- lm(log.val~Year+Time.Sample, data=all.birbs.PT) # model for 2-way anova

Anova(plasmaTSAM.fit, type="II") # generates a typical summary of the output



resid.anova.PT.all=resid(plasmaTSAM.fit) # generating residuals

qqnorm(resid.anova.PT.all) # plotting residuals for visual test of normality
qqline(resid.anova.PT.all)
shapiro.test(resid.anova.PT.all)
hist(resid.anova.PT.all)
boxplot(resid.anova.PT.all~all.birbs.PT$Time.Sample)

leveneTest (resid.anova.PT.all~all.birbs.PT$Time.Sample) #HoV test, normality will follow after once we get residuals
cochran.test(plasmaTSAM.fit, data = all.birbs.PT)
```

```{r}

all.birbs$Year<- as.factor(all.birbs$Year)

all.birbs$Size.Adjusted.Mass<-(all.birbs$Weight..g./(all.birbs$Wing..mm.^3))*10000

weight.birds<-all.birbs %>%
  filter(Weight..g.>0)
```

test w/ Size Adjusted Mass as Dependent Var:
```{r}

weight.fit<- lm(Size.Adjusted.Mass~Year+Time.Sample, data=weight.birds) # model for 2-way anova

Anova(weight.fit, type="II") # generates a typical summary of the output


resid.weight=resid(weight.fit) # generating residuals

qqnorm(resid.weight) # plotting residuals for visual test of normality
qqline(resid.weight)
shapiro.test(resid.weight)
hist(resid.weight)
boxplot(resid.weight~weight.birds$Time.Sample)

boxplot(weight.birds$Size.Adjusted.Mass~weight.birds$Time.Sample+weight.birds$Year)

cochran.test(weight.fit, data = weight.birds)
```


Behavior Analyses:

```{r}
feedcombo<-read.csv("Brazil Scan Sample 19_20combined.csv")
```

```{r}
feedcombo$logitfeeding<-logit(feedcombo$Feeding)
```

PERMANOVA:
```{r} 
behaviors<-feedcombo %>%
  dplyr::select(Rest, Preen, Locomotion, Peck, Probe, Skim)

behaviors19<-feedcombo %>%
  dplyr::filter(Year==2019) %>%
  dplyr::select(Rest, Preen, Locomotion, Peck, Probe, Skim)

feedcombo19<-feedcombo %>%
  dplyr::filter(Year==2019)

behaviors20<-feedcombo %>%
  dplyr::filter(Year==2020) %>%
  dplyr::select(Rest, Preen, Locomotion, Peck, Probe, Skim)

feedcombo20<-feedcombo %>%
  dplyr::filter(Year==2020)


vegan::adonis(behaviors~ Site*Year, data= feedcombo,  permutations=999, na.rm=TRUE)
adonis(behaviors20 ~ Site, data= feedcombo20,  permutations=999)
adonis(behaviors19 ~ Site, data= feedcombo19,  permutations=999)
```

canonical analyses and permdisp:
```{r}
canonanalysis1<- simper(behaviors, feedcombo$Site)
summary(canonanalysis1, ordered = TRUE, permutations=999)

canonanalysis2<- simper(behaviors19, feedcombo19$Site)
summary(canonanalysis2, ordered = TRUE, permutations=999)

canonanalysis3<- simper(behaviors20, feedcombo20$Site)
summary(canonanalysis3, ordered = TRUE, permutations=999)


dist.beh <- vegdist(behaviors) #creates bray-curtis values for the species across all observations
test1<- anova(betadisper(dist.beh,feedcombo$Site)) #now looking at the differences in the distance values for sites
test1

```

linear model-logit feeding:
```{r}
feedcombo$Year<-as.factor(feedcombo$Year)


feed.mod<-lm(logitfeeding ~ Year+Site+doy, data=feedcombo)
Anova(feed.mod)

cochran.test(feed.mod, data = feedcombo)
leveneTest(feedcombo$asinfeeding~feedcombo$Site)

resid.anovasfeed=resid(feed.mod) # generating residuals

qqnorm(resid.anovasfeed) # plotting residuals for visual test of normality
qqline(resid.anovasfeed)
shapiro.test(resid.anovasfeed)
hist(resid.anovasfeed)
boxplot(resid.anovasfeed~feedcombo$Site)

emmeans(feed.mod, pairwise ~ Site)
plot(asinfeeding~doy, data=feedcombo)
```

feeding plot:
```{r}
feedcombo$Year<-as.factor(feedcombo$Year)

feedingplotregcolor <- ggplot(feedcombo, aes(x=Site, y=Feeding, fill=Year)) +
  geom_boxplot() +
  theme_classic() +
  scale_x_discrete(limits=c("ISAL","ITFL"), labels=c("Salinas","Tidal Flats")) +
  scale_fill_brewer(labels=c("2019","2020"),palette = "Paired") +
  labs(y="Proportion of Time Spent Foraging")

feedingplotregcolor
```

NMDS PLots:
```{r}
behaviorall19<-feedcombo19 %>%
  dplyr::select(Peck,Probe,Skim,Rest,Preen,Locomotion)


NMDS<- metaMDS(behaviorall19)
```

```{r}
feedcombo19$Site<-as.factor(feedcombo19$Site)
locations<-data.frame(feedcombo19[,1])
names(locations)=""
locs1<-envfit(NMDS, env=locations)
behs1 <-envfit(NMDS, behaviorall19)
par(mfrow=c(1,1))
plot(NMDS, type="n")
points(NMDS,display="sites", cex = 0.8, pch=21, col="black", bg="orange")
#text(NMDS, display = "spec", cex=0.7, col="blue")
plot(locs1)
plot(behs1, col="black", cex=0.8)
ordihull(NMDS,groups= locations$Site,draw="lines",lty=3, col="gray60", label=F)
```

prep for ggplot:

```{r}
bird.NMDS.data<-locations
bird.NMDS.data$NMDS1<-NMDS$points[ ,1] #this puts the NMDS scores for the plots into a new dataframe. you could put them into an existing one if you preferred.
bird.NMDS.data$NMDS2<-NMDS$points[ ,2] 

colnames(bird.NMDS.data)<- c("Site","NMDS1","NMDS2")

stems<-colSums(behaviorall19) #total abundances for each species
spps <- data.frame(vegan::scores(NMDS, display = "species")) #dataframe of species scores for plotting
spps$species <- row.names(spps) # making a column with species names
spps$colsums <- stems #adding the colSums from above
spps.colmedian <- median(spps$colsums) #create an object that is the median of the abundance of the measured species
spps.colmean <- mean(spps$colsums) #creates a mean instead if you wish to use
spps2 <- subset(spps,spps$colsums > spps.colmean) #select the most abundant species. Could discard fewer by going something like - spps$colsums>(spps.colmedian/2) instead
spps2$species <- factor(spps2$species) #otherwise factor doesn't drop unused levels and it will throw an error
 
env.scores.bird <- as.data.frame(vegan::scores(locs1, display = "factors")) #extracts relevant scores from envifit
env.scores.bird <- cbind(env.scores.bird, env.variables = rownames(env.scores.bird)) #and then gives them their names

env.scores.beh <- as.data.frame(vegan::scores(behs1, display = "vectors")) #extracts relevant scores from envifit
env.scores.beh <- cbind(env.scores.beh, env.variables = rownames(env.scores.beh)) #and then gives them their names

#data for ellipse, in this case using the management factor
df_ell.bird.Site <- data.frame() #sets up a data frame before running the function.

veganCovEllipse <- function (cov, center = c(0, 0), scale = 1, npoints = 100) 
{
  theta <- (0:npoints) * 2 * pi/npoints
  Circle <- cbind(cos(theta), sin(theta))
  t(center + scale * t(Circle %*% chol(cov)))
}


for(ITFL in levels(bird.NMDS.data$Site)){
  df_ell.bird.Site <- rbind(df_ell.bird.Site, cbind(as.data.frame(with(bird.NMDS.data [bird.NMDS.data$Site==ITFL,],
                                                         veganCovEllipse(cov.wt(cbind(NMDS1,NMDS2),wt=rep(1/length(NMDS1),length(NMDS1)))$cov,center=c(mean(NMDS1),mean(NMDS2))))),Site=ITFL))
}
 

 
# data for labelling the ellipse
NMDS.mean.bird=aggregate(bird.NMDS.data[ ,c("NMDS1", "NMDS2")], 
                         list(group = bird.NMDS.data$Site), mean)
 
# data for labelling the ellipse
NMDS.mean=aggregate(bird.NMDS.data[,c("NMDS1", "NMDS2")], 
                    list(group = bird.NMDS.data$Site), mean)
 
```

ggplot:
```{r}
mult <- 2 #multiplier for the arrows and text for envfit below. You can change this and then rerun the plot command.
levels(bird.NMDS.data$Site) <- c("Salinas", "Tidal Flats")
levels(df_ell.bird.Site$Site) <- c("Salinas", "Tidal Flats")
df_ell.bird.Site[df_ell.bird.Site == "ISAL"] <- "Salinas"
df_ell.bird.Site[df_ell.bird.Site == "ITFL"] <- "Tidal Flats"
bird.nmds.gg1 <- ggplot(data = bird.NMDS.data, aes(y = NMDS2, x = NMDS1))+ #sets up the plot. brackets around the entire thing to make it draw automatically
    geom_path(data = df_ell.bird.Site, aes(x = NMDS1, y = NMDS2, group = Site, color=Site))+ #this is the ellipse, seperate ones by Site. If you didn't change the "alpha" (the shade) then you need to keep the "group 
    scale_alpha_manual(guide = FALSE,values=c(0.3, 0.5, 0.7, 0.9))+ #sets the shade for the ellipse
    geom_point(aes(color=Site), size = 3) + #puts the site points in from the ordination, shape determined by site, size refers to size of point
    #geom_text(data=spps2, aes(x=spps2$NMDS1, y=spps2$NMDS2, label=species), size = 3.3, hjust=1.1)+ #labelling the species. hjust used to shift them slightly from their points
    geom_segment(data = env.scores.beh,
                 aes(x = 0, xend = mult*NMDS1, y = 0, yend = mult*NMDS2),
                 arrow = arrow(length = unit(0.25, "cm")), colour = "grey") + #arrows for envfit.  doubled the length for similarity to the plot() function. NB check ?envfit regarding arrow length if not familiar with lengths
    geom_text(data = env.scores.beh, #labels the environmental variable arrows * "mult" as for the arrows
              aes(x = mult*NMDS1, y = mult*NMDS2, label=env.variables),
              size = 5,
              hjust = .5)+
    #geom_point(data=spps2, alpha = .6, shape = 4)+ #these are the species points, made lighter and a specific shape
    #scale_shape_manual(values = c(16,16))+ #sets the shape of the plot points instead of using whatever ggplot2 automatically provides
    
    coord_cartesian(xlim = c(-2,2))+  ## NB this changes the visible area of the plot only (this is a good thing, apparently). Can also specify ylim. Here in case you want to set xaxis manually.
    theme_classic()
bird.nmds.gg1.final<-bird.nmds.gg1+theme(axis.title = element_text(size = rel(1.5)), axis.text = element_text(color="black", size = 15)) + theme(axis.line = element_line(color="black", size =1)) + theme(legend.title=element_text(size=15), 
    legend.text=element_text(size=15))+ scale_color_manual(values=c("goldenrod1","dodgerblue"))
 
bird.nmds.gg1.final
```

2020:
```{r}
behaviorall20<-feedcombo20 %>%
  dplyr::select(Peck,Probe,Skim,Rest,Preen,Locomotion)


NMDS1<- metaMDS(behaviorall20)
```

```{r}
feedcombo20$Site<-as.factor(feedcombo20$Site)
locations<-data.frame(feedcombo20[,1])
names(locations)=""
locs1<-envfit(NMDS1, env=locations)
behs1 <-envfit(NMDS1, behaviorall20)
par(mfrow=c(1,1))
plot(NMDS, type="n")
points(NMDS,display="sites", cex = 0.8, pch=21, col="black", bg="orange")
#text(NMDS, display = "spec", cex=0.7, col="blue")
plot(locs1)
plot(behs1, col="black", cex=0.8)
ordihull(NMDS,groups= locations$Site,draw="lines",lty=3, col="gray60", label=F)
```

```{r}
bird.NMDS.data<-locations
bird.NMDS.data$NMDS1<-NMDS$points[ ,1] #this puts the NMDS scores for the plots into a new dataframe. you could put them into an existing one if you preferred.
bird.NMDS.data$NMDS2<-NMDS$points[ ,2] 

colnames(bird.NMDS.data)<- c("Site","NMDS1","NMDS2")

stems<-colSums(behaviorall20) #total abundances for each species
spps <- data.frame(vegan::scores(NMDS, display = "species")) #dataframe of species scores for plotting
spps$species <- row.names(spps) # making a column with species names
spps$colsums <- stems #adding the colSums from above
spps.colmedian <- median(spps$colsums) #create an object that is the median of the abundance of the measured species
spps.colmean <- mean(spps$colsums) #creates a mean instead if you wish to use
spps2 <- subset(spps,spps$colsums > spps.colmean) #select the most abundant species. Could discard fewer by going something like - spps$colsums>(spps.colmedian/2) instead
spps2$species <- factor(spps2$species) #otherwise factor doesn't drop unused levels and it will throw an error
 
env.scores.bird <- as.data.frame(vegan::scores(locs1, display = "factors")) #extracts relevant scores from envifit
env.scores.bird <- cbind(env.scores.bird, env.variables = rownames(env.scores.bird)) #and then gives them their names

env.scores.beh <- as.data.frame(vegan::scores(behs1, display = "vectors")) #extracts relevant scores from envifit
env.scores.beh <- cbind(env.scores.beh, env.variables = rownames(env.scores.beh)) #and then gives them their names

#data for ellipse, in this case using the management factor
df_ell.bird.Site <- data.frame() #sets up a data frame before running the function.

veganCovEllipse <- function (cov, center = c(0, 0), scale = 1, npoints = 100) 
{
  theta <- (0:npoints) * 2 * pi/npoints
  Circle <- cbind(cos(theta), sin(theta))
  t(center + scale * t(Circle %*% chol(cov)))
}


for(ITFL in levels(bird.NMDS.data$Site)){
  df_ell.bird.Site <- rbind(df_ell.bird.Site, cbind(as.data.frame(with(bird.NMDS.data [bird.NMDS.data$Site==ITFL,],
                                                         veganCovEllipse(cov.wt(cbind(NMDS1,NMDS2),wt=rep(1/length(NMDS1),length(NMDS1)))$cov,center=c(mean(NMDS1),mean(NMDS2))))),Site=ITFL))
}
 

 
# data for labelling the ellipse
NMDS.mean.bird=aggregate(bird.NMDS.data[ ,c("NMDS1", "NMDS2")], 
                         list(group = bird.NMDS.data$Site), mean)
 
# data for labelling the ellipse
NMDS.mean=aggregate(bird.NMDS.data[,c("NMDS1", "NMDS2")], 
                    list(group = bird.NMDS.data$Site), mean)
 
```

ggplot:
```{r}
mult <- 2 #multiplier for the arrows and text for envfit below. You can change this and then rerun the plot command.
levels(bird.NMDS.data$Site) <- c("Salinas", "Tidal Flats")
levels(df_ell.bird.Site$Site) <- c("Salinas", "Tidal Flats")
df_ell.bird.Site[df_ell.bird.Site == "ISAL"] <- "Salinas"
df_ell.bird.Site[df_ell.bird.Site == "ITFL"] <- "Tidal Flats"
bird.nmds.gg2<- ggplot(data = bird.NMDS.data, aes(y = NMDS2, x = NMDS1))+ #sets up the plot. brackets around the entire thing to make it draw automatically
    geom_path(data = df_ell.bird.Site, aes(x = NMDS1, y = NMDS2, group = Site, color=Site))+ #this is the ellipse, seperate ones by Site. If you didn't change the "alpha" (the shade) then you need to keep the "group 
    scale_alpha_manual(guide = FALSE,values=c(0.3, 0.5, 0.7, 0.9))+ #sets the shade for the ellipse
    geom_point(aes(color=Site), size = 3) + #puts the site points in from the ordination, shape determined by site, size refers to size of point
    #geom_text(data=spps2, aes(x=spps2$NMDS1, y=spps2$NMDS2, label=species), size = 3.3, hjust=1.1)+ #labelling the species. hjust used to shift them slightly from their points
    geom_segment(data = env.scores.beh,
                 aes(x = 0, xend = mult*NMDS1, y = 0, yend = mult*NMDS2),
                 arrow = arrow(length = unit(0.25, "cm")), colour = "grey") + #arrows for envfit.  doubled the length for similarity to the plot() function. NB check ?envfit regarding arrow length if not familiar with lengths
    geom_text(data = env.scores.beh, #labels the environmental variable arrows * "mult" as for the arrows
              aes(x = mult*NMDS1, y = mult*NMDS2, label=env.variables),
              size = 5,
              hjust = .5)+
    #geom_point(data=spps2, alpha = .6, shape = 4)+ #these are the species points, made lighter and a specific shape
    #scale_shape_manual(values = c(16,16))+ #sets the shape of the plot points instead of using whatever ggplot2 automatically provides
    
    coord_cartesian(xlim = c(-2,2))+  ## NB this changes the visible area of the plot only (this is a good thing, apparently). Can also specify ylim. Here in case you want to set xaxis manually.
    theme_classic()
bird.nmds.gg2.final<-bird.nmds.gg2+theme(axis.title = element_text(size = rel(1.5)), axis.text = element_text(color="black", size = 15)) + theme(axis.line = element_line(color="black", size =1)) + theme(legend.title=element_text(size=15), 
    legend.text=element_text(size=15))+ scale_color_manual(values=c("goldenrod1","dodgerblue"))
 
bird.nmds.gg2.final
```

combined plot:
```{r}
library(ggpubr)

combinedNMDS <- ggarrange(bird.nmds.gg1.final, bird.nmds.gg2.final,
                    labels = c("2019", "2020"),
                    hjust = -1.5,
                    ncol = 1, 
                    common.legend = TRUE ,
                    legend = "right")
combinedNMDS + theme_classic(base_size = 12)
```



